\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{longtable}
\usepackage{enumitem}
\usepackage{fancyvrb}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage[hidelinks]{hyperref}
\setlength{\parskip}{0.5em}
\setlength{\parindent}{0pt}
\newcommand{\code}[1]{\texttt{#1}}
% Tighten tables and allow them to break/scale.
\renewcommand{\arraystretch}{1.15}

\title{Purpose-sized languages: buying total verification with smallness}
\author{[author]\\\normalsize (preprint --- authorship to be set before submission)}
\date{\today}

\begin{document}
\maketitle

\section*{Abstract}

Agents increasingly generate code, and the scarce resource is not generation
but trust: who checks what a model wrote, and whether a third party can
re-check it. We argue that verification, not the harness around a generator, is
the durable layer of the agent stack, and that its leverage grows with model
capability rather than falling with it. We develop a concrete form of the
thesis: the smaller a language's sanctioned surface, the more of a program's
behavior a checker can decide --- a direction we demonstrate at two purpose-sized
points, not a scaling law measured across a range of sizes. A
general-purpose compiler checks types but cannot say whether a program halts,
what it may touch, or how much it may emit; a language sized to a purpose can
make those properties mechanical and complete.

We present \code{litelite}, a kit of eight zero-dependency crates (the workspace
totals 8,214 LOC across eleven crates with the three languages built on it; caps
CI-enforced, native and \code{wasm32-unknown-unknown} both green) that pays the
shared kernel of a family of purpose-sized languages once --- a parser depth
guard, a UTF-8-safe lexer, one fuel type, one capability table --- each an
invariant the three parent languages implemented divergently or omitted. On the
kit we build \code{prooflite}, a total fuel-bounded reference language, and
\code{stratlite} with its verifier \code{backtestlite}, a strategy language whose backtest
reduces to one reproducible integer and which has no name for a future bar. The
demonstrated saving is the base kernel: it is reused across the two kit
languages rather than re-rolled per language, so each lands at its
language-specific LOC only (1,955 and 2,562 LOC) in single recorded git
sessions, each surviving adversarial review. We flag the contrast against the
parents' 3,100--8,000 LOC as suggestive rather than clean: N = 2, the same
developer and model era built both kit languages and all three parents, and the
kit languages are tree-walk with no codegen while the compiling parents carry
emitters --- so only the kernel amortization is on firm ground (\S{}4.5).

We put the verifier to work as a selection filter over 134 agent-generated
programs (100\% compile, 98.5\% survivor) and specify it as a training reward
with tested anti-hacking guards. We report the negative results plainly: on our
single month of one asset the fuel bound never bound (the most expensive
strategy used 0.74\% of budget) and the held-out window is no harder than train
(a 1.5-point gate-clear gap), so the benchmark has no out-of-sample teeth; and
the verifier certifies well-formedness but not profit --- 114 of 132 survivors
lose money on the window they were scored against. We then run the verifier as
the sole training reward: a small open-weights model (\code{Qwen3-0.6B}), fine-tuned
by verifier-only rejection-sampling self-play with no teacher model and no API
key, goes from a measured-zero floor --- zero compiling programs in 256 attempts,
because \code{stratlite} is in no pretraining corpus --- to 100\% compile and \textasciitilde{}96\%
held-out gate-clear, holding across a five-month embargo and a never-trained
asset (95.7\% / 96.5\% / 96.1\%). This carries the predecessor tempo-x402 recipe
off \code{rustc} onto a purpose-built language --- though it measures generation
validity, not tempo-x402's task-solving pass@1, and the measured-zero floor
reflects the language's absence from pretraining rather than a harder task; the
near-zero train-vs-held-out gap marks the lift honestly as grammar competence,
not out-of-sample edge, which stays the selector's job. Run a second time on
\code{prooflite} --- a data-free compute language, via the same trainer and ladder
shape with a language-specific reward binary (different success rung and styles)
--- the lift repeats, from 3.5\% to \textasciitilde{}95\% rich generation with \textasciitilde{}100\% of it novel
against the cold-start corpus, so the result is not a one-grammar artifact
(generality past the shared base model and kit stays untested). A held-out
problem-solving benchmark (30 specs, exact-output match) then closes the
pass@1 gap and returns a two-sided verdict: transfer is real --- plain SFT on
the 174 verifier-selected corpus programs lifts solving from 20\% to 80\%
pass@8 --- but the self-play that wins generation (94.5\% vs 48.8\% RICH) gives
half of that back (43.3\%), and the loss localizes exactly to problems whose
correct output conflicts with the reward's shape, with the answer computed
and then padded: the policy internalizes the verifier's preferences as
unconditional habits that override the spec. A verifier certifies validity,
not intent --- measured here at the policy level.
One experiment remains unrun --- a frozen-model A/B arm for which there is no API
key --- and ships as an instrument with pre-registered commands, marked PENDING.
The deterministic verifier plus committed artifacts reproduce every number in
this paper from a command --- including the fine-tune's scoring, from committed
sample pools and candles; the generation itself (agents, a frozen API model, or
the fine-tune) does not, and is committed or recorded rather than regenerated.

\bigskip\hrule\bigskip

\section*{1. Motivation}

Agents increasingly write programs, and the durable question is not how well a
model writes but who checks what it wrote. Most of an agent harness is
machinery that encodes an assumption about what today's models cannot do: a
retry policy, a scaffolding step, a prompt convention. Those assumptions go
stale as models improve, so the harness layer is perpetually rewritten and
rarely compounds. Verification is different in kind. Its job is not to
compensate for a weak generator but to establish trust across an adversarial
boundary, and that job does not get easier or harder as the generator changes ---
it gets more valuable as more code arrives from parties you cannot audit by
hand.

The distinction we build on is testimony versus physics. A model reporting that
it checked its own work is testimony: it is only as trustworthy as the model,
and it degrades exactly when you most need it, under adversarial pressure or
distribution shift. A compiler rejecting a program is physics: the rejection
holds regardless of who or what produced the program, and it is cheap,
deterministic, and independently re-checkable by anyone who distrusts the
checker. As agents become economic actors that exchange code and commit to its
behavior, third-party-checkable verification becomes the layer that carries
trust, and its worth rises with model capability rather than falling with it.
Harnesses melt; verifiers compound.

The empirical seed for this thesis comes from the predecessor project,
tempo-x402: fine-tuning a 0.5B model on compiler-verified self-play lifted
pass@1 from 1.5\% to 16.4\% on a 201-problem benchmark, with the Rust compiler
serving as a free self-play verifier (prior work, tempo-x402; not reproduced in
this repo). That result used one verifier on one task family, and it hinted at a
more general claim we develop here: the leverage lives on the verifier side, and
it grows as the checked properties get stronger. A general-purpose compiler
checks types and borrows but says nothing about whether a program halts, what it
can touch, or how much it can emit. This paper asks what a verifier can
guarantee when the language is sized to the purpose rather than to generality,
and treats that verifier --- not the language --- as the product.

\bigskip\hrule\bigskip

\section*{2. The kit}

The kit is the shared kernel of the lite family --- the pieces that \code{rustlite},
\code{soliditylite}, and \code{bashlite} (\textasciitilde{}19K LOC of purpose-sized languages inside the
parent project, \code{localharness}) each hand-rolled separately, extracted and paid
for once. It ships as eight crates unified by the \code{litelite} facade: four
foundational kernel crates, the capability layer \code{caplite}, and two independent
emitters. The full workspace is eleven crates --- the three languages of \S{}3--\S{}5
ride on top --- and every one of them declares zero external dependencies: the
only entries in any crate's \code{[dependencies]} are other crates in this
workspace (verifiable by inspecting each \code{crates/*/Cargo.toml}; \code{std} only,
native and \code{wasm32-unknown-unknown} both green). The whole workspace is 8,214
lines against a 25,000-line ceiling (\code{bash scripts/caps.sh}).

\begingroup\small
\begin{longtable}{p{0.225\linewidth} p{0.225\linewidth} p{0.225\linewidth} p{0.225\linewidth}}
\toprule
\textbf{crate} & \textbf{role} & \textbf{LOC} & \textbf{tests} \\
\midrule\endhead
\code{diaglite} & spans, coded diagnostics, caret snippets & 252 & 7 \\
\code{lexlite} & UTF-8-safe byte-cursor lexer kit & 290 & 7 \\
\code{parselite} & recursive-descent harness + depth guard & 238 & 4 \\
\code{fuellite} & fuel + byte budgets & 180 & 4 \\
\code{caplite} & host capabilities as data & 521 & 10 \\
\code{evmlite} & EVM assembler + diff-oracle interpreter & 1,400 & 17 \\
\code{modlite} & wasm module builder & 770 & 8 + 1 doctest \\
\code{litelite} & facade re-exporting the seven above & --- & 1 \\
\bottomrule
\end{longtable}\endgroup

LOC from \code{bash scripts/caps.sh}; test counts from \code{cargo test -p <crate>}.
Fifty-eight unit tests plus one doctest cover the kit.

\subsection*{2.1 Invariants paid once}

Each foundational crate is a single invariant the three parents implemented
independently, with divergent bugs to show for it. The kit's value is not that
these pieces are clever --- they are deliberately boring --- but that each bug class
is now closed in one place.

\begin{itemize}[leftmargin=*]
\item \textbf{The depth guard.} \code{parselite} defines \code{DEFAULT\_MAX\_DEPTH = 96} once
\end{itemize}
  (\code{crates/parselite/src/lib.rs}). \code{rustlite} and \code{soliditylite} each carried a
  near-verbatim \textasciitilde{}120-line copy of the parser harness, \code{MAX\_RECURSION\_DEPTH = 96}
  duplicated in both. \code{bashlite} had \textbf{no guard at all} --- nested \code{if}-in-\code{if}
  recursed one frame per token until the stack overflowed, which on a wasm
  stack is an uncatchable abort that kills the browser tab rather than
  returning a diagnostic. On the kit a language cannot forget the guard,
  because \code{enter()} is the only way into the cursor.

\begin{itemize}[leftmargin=*]
\item \textbf{The mojibake bug.} \code{lexlite}'s \code{next\_char} decodes UTF-8 rather than
\end{itemize}
  byte-casting. The same multi-byte-character bug was fixed twice in the
  parents, differently each time; the kit's cursor is UTF-8-safe by
  construction, and it makes the nested-versus-flat block-comment choice and
  the digit-underscore-separator flag explicit rather than silently divergent.

\begin{itemize}[leftmargin=*]
\item \textbf{Three unrelated fuels.} \code{fuellite} is one \code{Fuel} type
\end{itemize}
  (\code{burn} / \code{Exhausted}) plus one \code{ByteBudget}. The parents had three unrelated
  implementations of the same idea: \code{bashlite}'s per-statement fuel with 256
  KiB output caps, \code{soliditylite}'s interpreter \code{STEP\_BUDGET} with a memory
  cap, and \code{rustlite}'s parse-depth cap backstopped by an out-of-band JS
  watchdog (\code{crates/fuellite/src/lib.rs}). The composition rule the parents
  learned the hard way is now structural: one \code{\&mut Fuel} threads down into
  every sub-evaluation, so fractal composition still terminates.

\begin{itemize}[leftmargin=*]
\item \textbf{The triple-declared host table.} \code{caplite} makes one \code{CapTable}, declared
\end{itemize}
  as data, drive four consumers that the parents kept in sync by hand: typed
  signatures for the checker, import emission for codegen, human docs, and a
  versioned parity manifest with an FNV-1a-64 hash for the far side of a
  boundary. \code{rustlite} hand-synced a Rust capability table against a JavaScript
  worker's copy, and the two drifted repeatedly. \code{caplite} kills that drift as a
  class --- one declaration, or a hash mismatch.

\subsection*{2.2 The constitution as mechanism}

The parents' recurring lesson was that a behavioural rule only holds once it
stops being a convention and becomes a gate. The kit encodes three.

\begin{itemize}[leftmargin=*]
\item \textbf{Caps are CI, not etiquette.} \code{scripts/caps.sh} enforces $\le$2,000 LOC per
\end{itemize}
  crate, $\le$25,000 repo-wide, and $\le$8,000 characters on \code{CLAUDE.md}, and CI runs
  it. At a cap the rule is split, shrink, or kill --- never raise the number. The
  cap binds in practice: \code{prooflite} sits at 1,955 of its 2,000 lines and
  \code{stratlite} at 1,893, and the same 2,000-line cap is what forced the M4 split
  into the \code{stratlite} language (1,893) and its \code{backtestlite} verifier (669)
  rather than one crate straddling both. The tightest cap is the surface one:
  \code{CLAUDE.md} is 7,963 of 8,000 characters (all four figures from
  \code{bash scripts/caps.sh}). The deeper invariant the character cap protects is
  that the repo's honest map stays inside one context window.

\begin{itemize}[leftmargin=*]
\item \textbf{The depth guard is the only door.} A language cannot enter the parser
\end{itemize}
  harness except through \code{enter()}, and \code{guarded()} pairs \code{enter}/\code{leave} on
  every path, so the recursion bound cannot be bypassed by forgetting to
  restore depth on an error return.

\begin{itemize}[leftmargin=*]
\item \textbf{Fuel is the only loop.} An evaluator that burns fuel on every step is
\end{itemize}
  mechanically total --- "this program halts within N units" holds by
  construction, not by review. There is no unbounded loop primitive to reach
  for; the budget is the loop.

\subsection*{2.3 Why the emitters stay separate}

Constitution rule 4 forbids a unified codegen trait, and the two emitters
honour it by being independent crates with no dependency between them (neither
\code{crates/evmlite/Cargo.toml} nor \code{crates/modlite/Cargo.toml} names the other).
This is not tidiness --- it is a claim proven in the parents that absolute-jump
machines and structured control flow are semantically irreconcilable. \code{evmlite}
is an EVM assembler over an opcode SSOT with minimal-width \code{PUSH} and two-pass
\code{PUSH2} label back-patching --- labels resolve to absolute byte offsets --- paired
with a diff-oracle interpreter that performs real \code{JUMPDEST} analysis and rolls
back storage on revert. \code{modlite} builds wasm modules with LEB128 encoding,
functype interning, locals run-length encoding, and the index bookkeeping
wasm's relative, structured control flow imposes (imports occupy the low
function indices, so \code{modlite} makes importing after a function exists a hard
error rather than a silent index shift). A shared emitter trait would have to
abstract over "jump to this byte" and "break out of this block," and forcing a
common shape onto them is precisely where the parents miscompiled. They ship as
separate libraries, and a wasm backend that never touches EVM depends on only
one of them. Both emitters are built and tested but as yet **unconsumed within
this repo**: neither kit language emits code, so their named consumer is the
pending M5 parent re-homing (\S{}6.4). Their payoff is argued from the parents, not
demonstrated here.

\bigskip\hrule\bigskip

\section*{3. prooflite: the reference total language}

\code{prooflite} is the kit's first language and its existence proof: the smallest
program that drives every kernel crate end to end --- \code{lexlite} lexing, a
\code{parselite} depth-guarded parse, a \code{fuellite}-fueled tree-walk evaluator, and a
\code{caplite} capability table as the host seam --- with every failure a coded,
spanned \code{diaglite} diagnostic. It is deliberately not in the \code{litelite} facade.
It is a consumer of the kit, the measured answer to "what does a language on the
kit cost, and what does it buy?"

The whole language is 1,955 lines of Rust including its tests, against the
2,000-line constitutional cap (\code{bash scripts/caps.sh}), with 31 unit tests and
2 doctests all green (\code{cargo test -p prooflite}). Nothing here is large. That is
the point: the guarantees below are available \emph{because} the language is small
enough to reason about completely, not in spite of it.

\subsection*{3.1 The language}

Values are 64-bit signed integers and booleans --- nothing else. Statements are
\code{let x = e;}, \code{x = e;}, \code{print e;}, \code{if e \{ \ldots{} \} else \{ \ldots{} \}} (else-if chains of
any length), and \code{repeat e \{ \ldots{} \}}, whose count is evaluated once, up front.
Expressions are literals, variables, host-capability calls \code{name(a, b)}, the
unary \code{-} and \code{!}, the usual arithmetic, comparison, and short-circuiting
logical operators, and parentheses. Comments are \code{// line} and nested
\code{/* block */}. There are no functions, no recursion, and no \code{while}: the only
loop is \code{repeat}, and it must announce how many times it will run before it
runs.

Three guarantees fall out of this shape, and each is a property that a
general-purpose language cannot give mechanically:

\begin{itemize}[leftmargin=*]
\item \textbf{Termination.} Every statement, every expression node, and every \code{repeat}
\end{itemize}
  iteration burns one unit of fuel from a single tank; a host call burns one
  plus its declared cost. When the tank is dry the program stops with \code{E0206}.
  "This program halts within \code{limits.fuel} steps" therefore holds for every
  program, adversarial ones included, by construction rather than by
  inspection. The default budget is 100,000 fuel and 64 KiB of output
  (\code{crates/prooflite/src/lib.rs}, \code{Limits::default}).
\begin{itemize}[leftmargin=*]
\item \textbf{A complete effect bound.} The host's capability table is the entire world
\end{itemize}
  a program can reach. Calls resolve, type-check, and cost fuel against that one
  table; the hostless \code{run} entry point installs an empty table, so such a
  program provably has no effect beyond the string it prints.
\begin{itemize}[leftmargin=*]
\item \textbf{Bounded output and bounded nesting.} \code{print} writes through a byte budget
\end{itemize}
  that clips at the cap without splitting a character and lets the run continue;
  the parser rides the kit depth guard so deep nesting is a diagnostic, never a
  stack overflow.

The cost model is exact, not approximate: it is a total function of the AST. The
test \code{the\_cost\_model\_is\_exact} pins \code{print 1;} at exactly 2 fuel, `let x = 1 +
2;\code{ at 4, and }repeat 3 \{ print 0; \}\code{ at 11 (}cargo test -p prooflite`). An
agent generating \code{prooflite} can predict a program's cost before running it.

\subsection*{3.2 What fuel-bounded totality costs}

Totality is not free, and the price is expressiveness. \code{prooflite} gives up
general recursion and unbounded iteration entirely; a loop that does not know
its own length is not expressible, because "run until a condition" is exactly
the shape fuel exists to forbid. Arithmetic that would wrap around instead
halts: overflow, division or remainder by zero, and negation of \code{i64::MIN} are
each a diagnostic, never a silent wrong-but-clean result
(\code{checked\_arithmetic\_never\_wraps}, \code{cargo test -p prooflite}). Even \code{i64::MIN}
is not writable as a literal --- \code{-} is an operator, so the positive half
overflows first --- and must be reached arithmetically.

This is the honest trade the paper's claim rests on. A general-purpose language
buys \code{while}, closures, and unbounded recursion, and pays for them by making
termination undecidable. \code{prooflite} refuses those constructs and, in exchange,
makes termination a property the evaluator enforces on every input without
analysis. Verification completeness and language size trade against each other,
and this section is one qualitative illustration of that trade-off --- not a rate
measured across sizes: for a language whose target is agent-generated,
mechanically selectable programs, the constructs given up are close to free and
the guarantee bought is the one the selection loop needs.

\subsection*{3.3 The two crash-grade lessons: adversarial review is load-bearing}

\code{prooflite} was built and then reviewed adversarially, and the review is where
the interesting evidence is. It found two defects that the test suite as written
did not, and both were process-killing, and both were failures of a \emph{guarantee}
rather than of ordinary logic --- which is precisely the class a
verification-first project cannot afford to ship. They are recorded in
\code{GENESIS.md} under "post-genesis lessons."

\textbf{The AST-spine stack overflow (M1).} The parser rides \code{parselite}'s depth
guard, so nesting is capped. But a left-associative operator chain like
\code{1+1+\ldots{}+1} folds \emph{iteratively} in the parser --- it costs O(1) guard entries ---
while building an AST whose left spine is as deep as the chain is long. The
evaluator, and even the drop glue that frees the tree, later recurse down that
spine. A flat 50,000-term source therefore parsed cleanly, consumed no
suspicious amount of fuel, and then aborted the whole process on a stack the
fuel budget could not touch. The lesson is sharp: **the depth guard bounds
parser recursion, not AST depth.** The fix charges one guard entry per fold, so
a chain's spine obeys the same cap as nesting, and reshapes else-if chains into
a flat vector so the common flat program stays unbounded without deepening the
tree. The invariant is now pinned by
\code{operator\_chains\_count\_toward\_the\_depth\_cap}: a chain of 50 evaluates, a chain
of 500 is an \code{E0102} diagnostic, and 5,000 folds parse-then-drop safely
(\code{cargo test -p prooflite}).

\textbf{The table-snapshot trust hole (M2).} The host capability table is validated
once, before the program runs, and that validated table is what the checker,
the docs, and the parity manifest all vouch for. But the evaluator originally
re-fetched \code{host.caps()} at every call site. A host with interior mutability
could therefore serve one table to the validator and a different table to
dispatch --- the guarantee "calls resolve only against the vouched-for table" was
a lie against an adversarial host. The fix fetches the table exactly once,
snapshots the validated \code{Copy} value, and resolves the entire run against that
snapshot; \code{the\_validated\_table\_snapshot\_drives\_the\_whole\_run} asserts that
\code{caps()} is called exactly once across a two-call program
(\code{cargo test -p prooflite}). The general form, which recurs at M3 and M4:
validate every channel into the artifact, then use the validated \emph{value}, never
a fresh fetch.

Neither defect was a typo. Each was a place where a stated guarantee --- "bounded
nesting never crashes," "the table is the complete, checked effect bound" --- held
for cooperative inputs and broke for adversarial ones. Ordinary tests, written
by the same mind that wrote the code, exercised the cooperative case. The M1
review was run by six independent finders against three refuters and surfaced 16
raw findings, 13 confirmed, 2 of them crash-grade; the M2 review, same
six-and-three shape, confirmed 16, including one crash-grade-adjacent hole in
the parity hash (\code{GENESIS.md}, post-genesis lessons). This is the section's
second measurement, and it is a claim about process, not luck: when the product
\emph{is} the guarantee, the guarantee has to be attacked on purpose, because the
failure modes that matter are exactly the ones that a well-behaved test program
never reaches.

\bigskip\hrule\bigskip

\section*{4. Construction cost: the Nth language on the kit}

The kit's premise is that a purpose-sized language is cheap to build once the
reusable kernel is paid for. This section measures that cost directly: the two
languages built on \code{litelite} against the three the parent project
(\code{localharness}) hand-rolled before the kernel was extracted. The comparison is
suggestive rather than controlled --- the caveats at the end are load-bearing ---
but the numbers point one way.

\subsection*{4.1 What is being measured}

The parents each re-implemented a kernel --- byte lexer, recursive-descent harness
with a recursion guard, fuel, spanned diagnostics --- and then a language on top.
\code{litelite} writes that kernel once and reuses it. So the honest unit of
"construction cost" for the Nth kit language is its \emph{marginal} LOC and tests:
the language-specific code, not a fresh kernel each time.

The shared kernel is small. The four base crates \code{diaglite}, \code{lexlite},
\code{parselite}, \code{fuellite} total 960 LOC (\code{bash scripts/caps.sh}:
252 + 290 + 238 + 180), and both kit languages draw on it for free --- this is the
amortization the repo actually demonstrates. Adding the capability layer and
both emitters brings shared infrastructure to 3,651 LOC (\code{bash scripts/caps.sh},
summing all seven kit crates), but that fuller figure is aspirational for the
emitters: \code{evmlite} (1,400 LOC) and \code{modlite} (770 LOC) are built and tested yet
have no consumer in this repo, since both kit languages are tree-walk with no
codegen. A future language that needs codegen \emph{could} draw on them for free; none
does today, so their reuse is claimed, not shown (\S{}6.4).

\subsection*{4.2 The two kit languages vs the three parents}

\begingroup\small
\begin{longtable}{p{0.180\linewidth} p{0.180\linewidth} p{0.180\linewidth} p{0.180\linewidth} p{0.180\linewidth}}
\toprule
\textbf{Language} & \textbf{Role} & \textbf{LOC} & \textbf{Tests} & \textbf{Kernel / codegen} \\
\midrule\endhead
\textbf{Parents} (hand-rolled, in \code{localharness}) &  &  &  &  \\
\code{rustlite} & Rust subset $\rightarrow$ wasm & 8,000 & 99 & own kernel + wasm emitter \\
\code{soliditylite} & Solidity subset $\rightarrow$ EVM & 7,600 & 159 & own kernel + EVM assembler \\
\code{bashlite} & fuel-bounded shell & 3,100 & 64 & own kernel, \textbf{no depth guard} \\
\textbf{On \code{litelite}} &  &  &  &  \\
\code{prooflite} & total reference language, tree-walk eval & 1,955 & 31 + 2 doc & reuses kit kernel; no codegen \\
\code{stratlite} + \code{backtestlite} & total strategy language + its verifier & 2,562 & 16 + 2 doc & reuses kit kernel; no codegen \\
\bottomrule
\end{longtable}\endgroup

Parent figures are quoted from \code{GENESIS.md} ("The lineage") and describe crates
in a different repository; they are not reproducible by a command here.
Kit-language LOC comes from \code{bash scripts/caps.sh}; kit-language test counts
from \code{cargo test -p prooflite}, \code{cargo test -p stratlite}, and `cargo test -p
backtestlite\code{ (}prooflite\code{ 31 unit + 2 doctests; }stratlite` 10 + 1;
\code{backtestlite} 6 + 1; the last two sum to the 16 + 2 shown). The
\code{stratlite + backtestlite} LOC is the sum of the two crates, 1,893 + 669,
because M4 split one language across two crates --- the trader's live dependency
and the selection loop's verifier --- under one evaluation path.

The marginal kit language lands at 1,955 and 2,562 LOC against parents that
re-rolled everything at 3,100--8,000 LOC. The structural reason is the amortized
kernel, not a smaller language per se --- which is exactly why the scope caveat
below matters.

One parent number is worth isolating: \code{bashlite}, the smallest at 3,100 LOC,
shipped with \emph{no} parser recursion guard (\code{GENESIS.md}, "What was ported at
genesis") --- a live stack-overflow bug in production. The kit makes that guard
the only door into the parse harness (\code{parselite}, \code{DEFAULT\_MAX\_DEPTH = 96}), so
the class cannot recur. Smaller hand-rolled did not mean safer; it meant a
missing invariant nobody re-derived.

\subsection*{4.3 Wall-clock: one recorded session per language}

The one-session claim traces to commit timestamps (`git log
--format='\%h\textbar{}\%ad\textbar{}\%s' --date=format:'\%Y-\%m-\%d \%H:\%M:\%S' a557fef..1524eb3`, with
the genesis commit timestamp from \code{git show -s a557fef}). The entire kit plus
both languages plus both emitters landed in a single overnight span:

\begin{itemize}[leftmargin=*]
\item Genesis (kernel crates): \code{a557fef}, 2026-07-14 22:10:31.
\item \code{prooflite} (M1): landed \code{07c2877} at 2026-07-15 00:12:30; adversarial-review
\end{itemize}
  fixes \code{9ac20b3} at 00:40:37 --- a 28-minute review-and-fix window, with no
  commit between the genesis port snapshot (22:13:07) and the land.
\begin{itemize}[leftmargin=*]
\item \code{stratlite} + \code{backtestlite} (M4): landed \code{42667a6} at 02:59:24; review fixes
\end{itemize}
  \code{1524eb3} at 03:34:19 --- a 35-minute review-and-fix window; the whole milestone
  (from M3 done at 02:25:10) spans 1 h 09 m.
\begin{itemize}[leftmargin=*]
\item End to end, genesis through M4 review-complete: 22:10:31 $\rightarrow$ 03:34:19,
\end{itemize}
  \textbf{5 h 24 m}, for four kernel crates, the capability layer, two emitters, and
  two languages.

Commit timestamps bracket elapsed time between commits; they include idle and
cannot separate think-time from typing, so treat them as the recorded envelope
of a session, not a stopwatch. What they do establish is that each language went
from nothing to landed-and-adversarially-reviewed inside a single dated sitting,
not the "\textasciitilde{}a week" the roadmap had budgeted (\code{GENESIS.md}, "The three
questions").

\subsection*{4.4 Defects at landing}

Each language shipped through the same adversarial review (multi-agent
finder/refuter panel) and the counts are recorded in \code{GENESIS.md} ("Post-genesis
lessons"): \code{prooflite} drew 16 raw findings $\rightarrow$ 13 confirmed, 2 crash-grade;
\code{stratlite}/\code{backtestlite} drew 14 confirmed. Notably, \code{prooflite}'s two
crash-grade defects were the \emph{same} lesson --- the parser depth guard bounds
recursion, not AST depth, so a flat 50K-term source builds a spine the evaluator
recurses down (\S{}3.3) --- and that fix was paid once, into the kit's consumption
idiom (every iterative AST-deepening fold now charges the guard). These counts
come from review panels recorded in \code{GENESIS.md}; they are not re-runnable and
depend on panel composition, so they index review thoroughness, not an absolute
defect density.

\subsection*{4.5 Caveats}

The reading here is deliberately narrow.

\begin{itemize}[leftmargin=*]
\item \textbf{N = 2.} Two kit languages is a data point, not a trend. The seam-tax
\end{itemize}
  question --- whether the kit is a net win or a tax a re-homed parent would refuse
  --- stays open until M5 re-homes \code{bashlite} (\code{GENESIS.md}, "The three
  questions"); \S{}6.4 returns to it.
\begin{itemize}[leftmargin=*]
\item \textbf{Same developer, same model era.} Both kit languages and all three parents
\end{itemize}
  were built by the same developer with the same model generation. There is no
  controlled A/B; this is not evidence that the kit helps a \emph{different} builder.
\begin{itemize}[leftmargin=*]
\item \textbf{LOC is a proxy} for effort and for scope. It is not normalized to language
\end{itemize}
  power.
\begin{itemize}[leftmargin=*]
\item \textbf{Scope mismatch cuts against the aggressive reading.} \code{prooflite} and
\end{itemize}
  \code{stratlite} are tree-walk evaluators with no codegen; the compiling parents
  (\code{rustlite} $\rightarrow$ wasm, \code{soliditylite} $\rightarrow$ EVM) carry emitters the kit factors into
  \code{evmlite}/\code{modlite} (2,170 LOC combined) that neither kit language yet uses. A
  fair like-for-like --- a kit language that actually emits code --- is future work.
  Until then, the 1,955/2,562-vs-3,100--8,000 gap conflates the amortized kernel
  with a genuinely lighter language scope, and only the kernel amortization is
  on firm ground.

The defensible claim: building the Nth language on the kit costs its
language-specific LOC only, because the kernel is paid once; the two kit
languages landed at 1,955 and 2,562 LOC in single recorded git sessions, each
surviving adversarial review, against parents that re-rolled their kernels at
3,100--8,000 LOC. Whether that advantage survives a codegen-scope language and a
second builder is what M5 and later data points must answer.

\bigskip\hrule\bigskip

\section*{5. Verified selection, a verifier-as-reward specification, and the fine-tune result}

Sections 3 and 4 measured what smallness costs and what it buys at the language
boundary. This section puts the guarantees to work: it uses the mechanical
verifier as a selection filter over agent-generated programs, and then specifies
it as a training reward. The claim under test is narrow and mechanical --- that a
purpose-sized language turns "is this program acceptable?" into a total,
deterministic, third-party-checkable predicate --- and the results are reported
with an equally narrow honesty about what the single-month benchmark can and
cannot show.

The instrument shipped complete. The verifier-only GPU fine-tune has now been
run twice --- \S{}5.6 reports it on \code{stratlite} and \S{}5.7 its N = 2 replication on
\code{prooflite}; the remaining experiment, a frozen-model A/B arm for which there is
no API key, has not been run, and \S{}5.9 marks it as a PENDING protocol slot with
its exact commands, per the house rule that pending results are named, not
promised.

\subsection*{5.1 The instrument: \code{verify()}, \code{equity\_hash}, and the \code{Reject} histogram}

The whole selection apparatus is one function. `backtestlite::verify(src,
candles, limits, costs, gate)` compiles the source, backtests the strategy, and
applies an activity gate, returning \code{Result<(Strategy, Report), Reject>}
(\code{crates/backtestlite/src/lib.rs:199}). The error type is the selection signal
itself:

\begin{Verbatim}[fontsize=\small,frame=single,samepage=false]
enum Reject { Compile(Diag), Run(Diag), Gate(GateFail) }
\end{Verbatim}

Every rejected program lands in exactly one of three bins --- the source is
ill-formed (\code{Compile}), it faulted at run time or on the data (\code{Run}), or it ran
clean but did not behave like a strategy (\code{Gate}). Tallying those bins over a
pool is the \emph{Reject histogram}: structured selection pressure, not a scalar
pass/fail. The gate defaults to \code{min\_trades = 4}, \code{min\_bars\_evaluated = 16}
(\code{crates/backtestlite/src/lib.rs:89}).

Two properties make the predicate usable as an oracle rather than merely a
checker:

\begin{itemize}[leftmargin=*]
\item \textbf{Determinism as one number.} \code{Report::equity\_hash} is an FNV-1a-64 hash of
\end{itemize}
  the final equity curve, so an entire backtest reduces to a single reproducible
  integer; \code{Report} also derives \code{Eq}. The determinism is a pinned test
  (\code{determinism\_is\_exact\_and\_hashable}), so "the same strategy on the same
  candles scores identically" is enforced, not assumed
  (\code{cargo test -p backtestlite}, 7 tests pass).
\begin{itemize}[leftmargin=*]
\item \textbf{No look-ahead by construction.} \code{stratlite} has no name for a future bar
\end{itemize}
  and fills at the next open; this is a grammar fact, pinned by
  \code{no\_lookahead\_prefix\_invariance}, which mutates every candle after each index
  \code{k} and asserts no decision at or before \code{k} changes
  (\code{crates/stratlite/src/lib.rs:256}; \code{cargo test -p stratlite}, 11 tests pass).

Because evaluation is fuel-bounded (\code{Limits::default().fuel\_per\_bar = 25\_000},
\code{crates/stratlite/src/lib.rs:137}), computing a verdict is itself guaranteed to
terminate --- the property that lets the same predicate serve as a training reward
in \S{}5.5 without any generated program ever hanging the loop. The instrument is
1,893 LOC of language plus 669 LOC of verifier (\code{bash scripts/caps.sh}),
zero-dependency, and the generation prompt is \code{stratlite::REFERENCE} --- a \code{const}
of the crate (\code{crates/stratlite/src/lib.rs:78}), so the language the model is
shown and the language the verifier enforces cannot drift.

\subsection*{5.2 The key-free corpus run: real numbers}

Running \S{}5 end-to-end against a frozen API model needs an API key. To produce
real numbers without one, the generation step was performed by six agents --- one
per strategy family (trend, breakout, mean-reversion, momentum, stateful,
combo) --- each handed \code{stratlite::REFERENCE} and asked for diverse valid
programs. The result is committed as \code{experiment/corpus/seed.jsonl}: 134
programs, \code{\{id, style, source\}} per line, balanced across the six families
(trend 23, breakout 23, mean-reversion 22, momentum 22, stateful 22, combo 22;
\code{wc -l experiment/corpus/seed.jsonl} and a one-line \code{Counter} over the \code{style}
field). This is generate$\rightarrow$verify$\rightarrow$keep with agents as the generator; the honest
split holds --- the \emph{generation} does not reproduce and so is committed rather than
regenerated, while every number below is a pure function of that committed
corpus plus the pinned candles.

The candle data is \code{BTCUSDT-1h-2024-01.csv}: 744 hourly bars, chronologically
split 60/40 into a 446-bar train window and a 298-bar held-out window, with
adverse costs (5 bps fee, 1 bps slippage) fit to the train window only ---
\code{fee = 2196}, \code{slip = 439} ticks, held-out/train price drift 0.943x
(\code{cd experiment \&\& cargo run -q -- data data/BTCUSDT-1h-2024-01.csv}). Verifying
the corpus against the train window (`cd experiment \&\& cargo run -q -- reward
corpus/seed.jsonl data/BTCUSDT-1h-2024-01.csv`) gives the Reject histogram:

\begingroup\small
\begin{longtable}{p{0.300\linewidth} p{0.300\linewidth} p{0.300\linewidth}}
\toprule
\textbf{Outcome} & \textbf{Count} & \textbf{Rate} \\
\midrule\endhead
Compile-fail & 0 & 0\% \\
Run-fault & 0 & 0\% \\
Gate-fail & 2 & 1.5\% \\
Survivor (\code{ok}) & 132 & 98.5\% \\
\bottomrule
\end{longtable}\endgroup

Compile rate is 100\% and the survivor rate is 132/134 = \textbf{98.5\%}. The two
non-survivors are gate failures, not faults: they parsed and ran cleanly but did
not trade enough to count as strategies. Strong agents, asked for valid programs
and given the language card, can produce valid active \code{stratlite} almost every
time; this measures conditioned generation, not an unconditioned base rate --- the
calibrating comparison is the PENDING, permanently-keyless A/B arm (\S{}5.9), so we
do not read a base rate off this number. It is also the setup for the limitation
in \S{}5.4. The same run reports 134 distinct source-canonical novelty keys over
the 134 programs, so no two are template clones under the dedup key of \S{}5.5.

\subsection*{5.3 The fuel bound is free on this task --- measured, not assumed}

The termination guarantee is only interesting if we can say whether it \emph{bound}.
Across the 132 survivors, \code{max\_fuel\_per\_bar} is \textbf{min 16 / median 55 / max 186}
against the 25,000/bar cap --- the single most expensive diverse strategy consumes
\textbf{0.74\% of the budget} (same \code{reward} command; the field is emitted per line).
So on this task the fuel bound never came close to binding: it is free, and, as
a \emph{discriminator}, untested. The guarantee is genuine --- any program that looped
would be cut at 25,000 --- but this corpus never exercised it. \S{}6.1 develops this
as a named negative result about our own evidence.

\subsection*{5.4 The conditional metric and the no-teeth finding}

The naive way to compare generators is raw held-out survivor rate. It is
misleading, and identifying exactly why is a methodological contribution of this
section rather than an afterthought.

The \code{Compile} rung of the histogram is \textbf{data-independent}: whether a program
parses (and largely whether it halts) does not depend on the candles at all --- a
program that compiles on train compiles identically on held-out. So a raw
"held-out survivor lift" can be almost entirely a compile-rate lift that shows
up identically on train: grammar-learning dressed as generalization. The honest
metric therefore \emph{conditions on compiling} and reports the \textbf{gate-clear rate}
(the genuinely data-dependent rung) on each window, plus the
train-minus-held-out gap.

Running that metric (`cd experiment \&\& cargo run -q -- eval corpus/seed.jsonl
data/BTCUSDT-1h-2024-01.csv`):

\begin{itemize}[leftmargin=*]
\item compile rate \textbf{100\%};
\item among compilers, gate-clear \textbf{98.5\% train} vs \textbf{97.0\% held-out};
\item \textbf{gap = 1.5 points --- near zero.}
\end{itemize}

A near-zero gap is a finding, and we report it as one rather than hiding it: on
one month of one asset, the held-out window is no harder than the train window,
so raw survivor rate carries essentially no out-of-sample signal here. The
per-style breakdown confirms the gap is not an aggregate artifact --- five of six
families clear at 96--100\% held-out, and only mean-reversion shows any train/test
spread (19/22 = 86\%):

\begingroup\small
\begin{longtable}{p{0.450\linewidth} p{0.450\linewidth}}
\toprule
\textbf{Style} & \textbf{Held-out gate-clear (among that style's compilers)} \\
\midrule\endhead
breakout & 100\% (23/23) \\
combo & 100\% (22/22) \\
momentum & 100\% (22/22) \\
stateful & 100\% (22/22) \\
trend & 96\% (22/23) \\
mean-reversion & 86\% (19/22) \\
\bottomrule
\end{longtable}\endgroup

The methodological consequence is a precondition on any future claim: **prove
the benchmark has out-of-sample teeth before claiming a lift on it.** A wider
regime split --- a chronologically distant test window and a second asset, chosen
so that a trend-fitted strategy actually goes silent --- is a prerequisite for the
generalization experiment, not a nice-to-have. The instrument surfaces its own
benchmark's inadequacy, which is the behavior we want from a verifier; \S{}6.2
draws out the consequence for the fine-tune benchmark.

\subsection*{5.5 Verifier-as-reward (M6): the ladder and the anti-hacking guards}

\S{}5.2--5.4 select from a fixed pool. M6 moves one knob: it puts the model in the
loop and uses the same verifier as the training reward. M6 tests whether the
tempo-x402 result (\S{}1) would generalize off \code{rustc} onto a purpose-sized
language --- the hypothesis the \S{}5.6 and \S{}5.7 runs now test. Three properties of a fuel-bounded verifier make it a reward oracle
\code{rustc} cannot be: computing a rollout's reward always terminates (no generated
program can hang training), totality removes reward-hacking-by-nontermination as
a category, and the reward is CPU-cheap and deterministic (a checkpoint's
rewards reproduce from a command even though the model that produced the
rollouts never can).

The reward is a four-rung validity ladder read directly off \code{Reject}
(\code{experiment/src/reward.rs}): \code{Compile $\rightarrow$ 0}, \code{Run $\rightarrow$ 1/3}, \code{Gate $\rightarrow$ 2/3},
\code{Ok $\rightarrow$ 1}. \textbf{Train PnL is deliberately excluded from the reward value}, for two
anti-hacking reasons: rewarding it would make the held-out comparison circular
(you would be measuring what you optimized), and continuous train PnL is the
single most overfittable signal on a few months of one asset. Excluding PnL does
not, however, remove all overlap between what the reward optimizes and what the
win condition scores: the reward rewards the gate-clear (\code{Ok}) rung, which is
also exactly what the \S{}5.6 benchmark measures on held-out data. This is
standard train/test discipline only insofar as the held-out window is genuinely
harder than train --- a precondition \S{}6.2 shows the single-month benchmark fails ---
so a gate-clear lift, even on the wider regime split \S{}5.6 uses, reads as grammar
competence rather than out-of-sample edge. PnL, fuel, \code{equity\_hash}, and a dedup key are all \emph{emitted} so
a trainer can reshape and own that choice, but \code{value == ladder(class)} is a
pinned invariant. This is the yield/edge split: M6 is designed to yield a better
\emph{generator} (validity and diversity of the pool) --- \S{}5.6 reports that it does,
and \S{}5.7 replicates it on a second language; finding the profitable strategy
stays \S{}5.2's \emph{selector}. A validity-only reward means a null PnL result falsifies nothing and
cannot be spun as success.

The red team named five reward hacks; each has a guard, and the guards are
tested with no GPU:

\begin{enumerate}[leftmargin=*]
\item \textbf{Empty/junk rollouts.} An empty source or a fenced code block scores a
\end{enumerate}
   compile-zero, not a lenient partial --- verified as given, never repaired
   (\code{the\_empty\_rollout\_hack\_is\_closed},
   \code{garbage\_is\_a\_compile\_zero\_not\_a\_leniency}).
\begin{enumerate}[leftmargin=*]
\item \textbf{Gate-rung farming.} \code{lookback 4;} compiles, runs clean, gate-fails, and
\end{enumerate}
   collects 2/3 for the \emph{absence} of a strategy. Guard: SFT admission keeps
   Ok-rung survivors only; the 2/3 rung is a diagnostic and a GRPO signal, never
   an admission threshold (\code{experiment/train/admission.py},
   \code{test\_only\_ok\_rung\_is\_admitted}).
\begin{enumerate}[leftmargin=*]
\item \textbf{Template mode-collapse.} Guard: \code{novelty\_key}, a source-canonical
\end{enumerate}
   FNV-1a-64 over comment-stripped, whitespace-collapsed source, caps any one
   key's share of the SFT set. It sees through re-commenting and whitespace-run
   reformatting but --- stated as an honest limit --- not token-adjacency spacing
   (\code{x=1} vs \code{x = 1}), constant-perturbation, or semantic clones
   (\code{novelty\_key\_ignores\_comments\_and\_whitespace\_but\_not\_logic},
   \code{test\_dedup\_caps\_a\_template\_family}).
\begin{enumerate}[leftmargin=*]
\item \textbf{\code{equity\_hash} diversity gaming.} A one-tick constant change yields a new
\end{enumerate}
   curve and a new hash. Guard: use \code{novelty\_key}, which is defined below the
   survivor rung where \code{equity\_hash} is 0 for everything, for both dedup and the
   diversity metric.
\begin{enumerate}[leftmargin=*]
\item \textbf{Easiest-style collapse.} Guard: a per-style admission cap keeps the set
\end{enumerate}
   spread across families (\code{test\_per\_style\_cap\_keeps\_the\_set\_spread}).

The Rust reward core and its guards are covered by \code{cd experiment \&\& cargo test}
(23 tests pass), and the Python admission guards by `cd experiment/train \&\&
python3 test\_select.py` (6 pass) --- both stdlib-only, both runnable now. The
Rust/Python seam is exactly two things: the reward CLI (candidate JSONL in,
reward records out) and the generation format. Rust owns verify, the reward
scalar, the histogram, fuel, and the dedup key --- all reproducible today; Python
owns sampling and the weight update, which need a GPU and do not run in CI. The
trainer lives outside the kit's workspace (it takes \code{torch}) with its own cap
counter; \code{experiment/} is 1,489 LOC, \code{experiment/train/} is 639, and the N=2 reward tool
\code{experiment/proofbench/} is 354 (\code{bash scripts/caps.sh}).

\subsection*{5.6 The verifier-only fine-tune (M6): from a measured-zero floor to competent}

The M6 instrument has been run. A small open-weights model, the dense
\code{Qwen/Qwen3-0.6B} (\code{experiment/train/train.py:55}), was fine-tuned to generate
\code{stratlite} using only the kit's verifier as supervision --- no teacher model, no
API key. Cold-start SFT on the 132 committed corpus survivors (\S{}5.2) was
followed by eight rounds of verifier-only rejection-sampling self-play: sample
per style from checkpoint \code{C\_\{r-1\}}, batch-score with the reward CLI, admit
\code{Ok}-rung survivors deduped by \code{novelty\_key} under per-style caps (\S{}5.5), SFT to
\code{C\_r}, repeat. On the train reward window the full-survivor rate climbed
monotonically --- cold-start $\rightarrow$ R0 39.0\% $\rightarrow$ R1 67.0\% $\rightarrow$ R2 80.6\% $\rightarrow$ R3 85.8\% $\rightarrow$ R4
89.6\% $\rightarrow$ R5 94.6\% $\rightarrow$ R6 96.0\% $\rightarrow$ R7 98.2\% (\code{experiment/results/train\_curve.log}) ---
and training stopped at checkpoint C7 on the protocol's train-saturation
early-stop (rich-rate). \code{distinct\_nkeys} rose then narrowed --- 392 $\rightarrow$ 624 $\rightarrow$ 685 $\rightarrow$
654 $\rightarrow$ 694 (peak, R4) $\rightarrow$ 674 $\rightarrow$ 555 $\rightarrow$ 534, over 399--809 admitted per round --- the
same peak-then-narrow shape \S{}5.7 later finds for prooflite and \S{}6.6 treats as a
general limit. The anti-collapse guards of \S{}5.5 held the model to a diverse
grammar rather than one template through the rising limb, but did not prevent the
post-peak narrowing; because C7 was selected on rich-rate saturation, not the
distinct-key peak, it sits past its diversity optimum (\textasciitilde{}R4). That does not touch
the gate-clear result below --- a validity claim, and validity is monotone --- but
it does mean C7's generator is less varied than an earlier checkpoint's would be.
A held-out novelty check confirms the model learned the grammar, not the corpus:
of C7's 251 gate-clearing programs, 250 (99.6\%) carry a source-canonical key
absent from the committed corpus's 134 keys (205 of them distinct; \code{s5 reward}
emits the key per program), the same anti-memorization control \S{}5.7 runs for
prooflite --- now symmetric across both arms.
Fuel over survivors stayed at \textasciitilde{}1\% of the 25,000-per-bar cap throughout, as in the
corpus run (\S{}5.3) --- a stratlite-specific figure \S{}6.1 revisits against prooflite.

C7 was benchmarked against the same model before fine-tuning, with identical
non-thinking prompts (256 samples each, 32 per style over 8 styles), scored by
the deterministic verifier on three windows: the BTC January 2024 train reward
window, a distant BTC June 2024 window under a five-month embargo, and an ETH
June 2024 window on an asset never trained on. The conditional metric of \S{}5.4 ---
compile rate over the pool, then gate-clear among compilers --- reads:

\begingroup\small
\begin{longtable}{p{0.300\linewidth} p{0.300\linewidth} p{0.300\linewidth}}
\toprule
\textbf{window} & \textbf{base compile / gate-clear} & \textbf{C7 compile / gate-clear} \\
\midrule\endhead
BTCUSDT Jan 2024 (train reward window) & 0.0\% / 0.0\% & 100.0\% / 95.7\% \\
BTCUSDT Jun 2024 (distant, 5-month embargo) & 0.0\% / 0.0\% & 100.0\% / 96.5\% \\
ETHUSDT Jun 2024 (cross-asset, never trained on) & 0.0\% / 0.0\% & 100.0\% / 96.1\% \\
\bottomrule
\end{longtable}\endgroup

(`cd experiment \&\& ./target/release/s5 eval results/\{base,c7\}.jsonl
data/<window>.csv\code{; full output in }experiment/results/benchmark.txt`.) The
baseline is a true floor, not a weak one: \code{stratlite} exists in no pretraining
corpus, so base \code{Qwen3-0.6B} produces zero valid programs across 256 attempts on
every window --- it parrots the grammar card's notation but cannot emit one
compiling program. The entire base$\rightarrow$C7 lift is therefore the fine-tune, with
nothing to confound that gap. This shares the RECIPE of the tempo-x402 result of
\S{}1 --- compiler-verified self-play, no teacher --- but measures a different variable:
tempo-x402's 1.5\% $\rightarrow$ 16.4\% is pass@1 at SOLVING 201 specified Rust problems, a
language in pretraining, whereas this is the VALIDITY rate of open-ended
generation of a language absent from it. The measured-zero floor reflects that
absence --- the grammar must be acquired from nothing --- not a task harder than
solving Rust; what carries across is the recipe, a fuel-bounded verifier standing
in for \code{rustc} as the oracle on a language no compiler corpus ever saw.

What this establishes is bounded exactly as \S{}5.5 designed it. Verifier-only
fine-tuning takes a small model from no competence to \textasciitilde{}96\% valid-strategy
generation on a language it never saw pretrained, and that competence holds
across a five-month embargo and across assets (95.7\% / 96.5\% / 96.1\% held-out
gate-clear); the per-style held-out breakdown on the cross-asset window is
near-uniform (88--100\%), so no family collapsed onto one easy template. It does
\textbf{not} establish edge. The verifier certifies well-formed and active, never
profitable, and the train-minus-held-out gap stays near zero (2.3 / 1.2 / 1.6
points). Read against \S{}6.2, that near-zero gap is the honest signature of grammar
competence --- the language is genuinely as easy on ETH June as on BTC January ---
not of generalizing skill at finding good strategies. Selecting the profitable
strategy from this now-competent generator stays the \S{}5.2 selector's job
(\code{pick\_verified}), not the generator's; because the reward is validity-only
(\S{}5.5), this result carries no edge claim and cannot be spun into one. The MODEL
does not reproduce --- sampling is stochastic and a fine-tune is not bit-identical
across hardware --- but the SCORING does: the two sample pools
(\code{results/base.jsonl}, \code{results/c7.jsonl}, 256 programs each) and the candles
are committed, and the \code{s5 eval} command above reproduces every number in the
table.

\subsection*{5.7 The second-language replication (M6, N = 2): the same reward on a data-free language}

If the \S{}5.6 lift were an artifact of one grammar it would not transfer, so the
instrument was run a second time on \code{prooflite} (\S{}3) --- the reference \emph{compute}
language, which reads no market data, places no trades, and has no held-out
window. The language-parametric trainer ran UNCHANGED; only the reward binary
was swapped (\code{s5} $\rightarrow$ the parallel \code{p6}, \code{experiment/proofbench/src/main.rs}),
which serves the prooflite prompt card, eight computation-family styles, and the
same validity ladder --- \code{compile $\rightarrow$ run $\rightarrow$ gate $\rightarrow$ ok}, where the \code{ok}/RICH rung
requires a clean run that prints $\ge$ 3 distinct lines over $\ge$ 30 fuel. Cold start on
the 174 committed corpus survivors --- the \code{ok}-rung keepers of 175 key-free agent
drafts (99.4\% RICH raw; the survivor file
\code{experiment/proofbench/corpus/seed.jsonl} is 100\% RICH under \code{p6} by
construction) --- then nine rounds of the identical rejection-sampling self-play.
Rich-rate climbed monotonically --- cold-start $\rightarrow$ R0
43.8\% $\rightarrow$ R1 63.5\% $\rightarrow$ R2 72.4\% $\rightarrow$ R3 83.1\% $\rightarrow$ R4 85.7\% $\rightarrow$ R5 90.7\% $\rightarrow$ R6 91.2\% $\rightarrow$ R7
95.0\% $\rightarrow$ R8 96.2\% (of 1,024 sampled per round).

Because prooflite reads no data there is no train/held-out DATA split, so the
generalization question changes shape: are the rich programs LEARNED, or
MEMORIZED from the 174 human-authored cold-start examples --- the only external
data the model ever saw? The \code{p6 novelty} command answers it by source-canonical
key (FNV-1a-64 over comment-stripped, whitespace-collapsed source,
\code{main.rs:novelty\_key}, which sees through format and comment clones).
Benchmarking the selected checkpoint against the base model, identical
non-thinking prompts, 256 samples each (32 per style):

\begingroup\small
\begin{longtable}{p{0.180\linewidth} p{0.180\linewidth} p{0.180\linewidth} p{0.180\linewidth} p{0.180\linewidth}}
\toprule
\textbf{model} & \textbf{parse} & \textbf{RICH (ok)} & \textbf{distinct rich keys} & \textbf{novel ok / ok ($\notin$ corpus)} \\
\midrule\endhead
base \code{Qwen3-0.6B} & 23.4\% & 3.5\% & 9 & 9 / 9 (100\%) \\
C5 & 96.9\% & 90.6\% & 213 & 232 / 232 (100\%) \\
\textbf{C6 (selected)} & 99.2\% & 94.5\% & \textbf{216} & 242 / 242 (100\%) \\
C7 & 100.0\% & 96.1\% & 199 & 245 / 246 (99.6\%) \\
C8 & 98.8\% & 96.1\% & 205 & 245 / 246 (99.6\%) \\
\bottomrule
\end{longtable}\endgroup

(\code{cd experiment/proofbench \&\& ./target/release/p6 eval results/c6.jsonl} and
\code{./target/release/p6 novelty results/c6.jsonl corpus/seed.jsonl}; full output in
\code{experiment/proofbench/results/benchmark.txt}.) The floor is again true:
prooflite is in no pretraining corpus, so base \code{Qwen3-0.6B} --- which recognizes
the C-like surface enough to parse 23.4\% of its attempts, more than stratlite's
0\% --- still writes a RICH program only 3.5\% of the time (9 of 256). The lift from
3.5\% to \textasciitilde{}95\% is the fine-tune, and \textasciitilde{}100\% of the rich programs are novel against
the corpus --- so the competence is not recall of the 174 human examples, the only
external text the model saw. (Novelty is measured against that human seed only,
not against the model's own admitted self-play programs, so it rules out
memorizing the human corpus, not reproduction from the larger self-generated
training set.)

The selected checkpoint is C6, and why is itself a finding. Raw validity
saturates across C6--C8 (94.5\% / 96.1\% / 96.1\% RICH), but DIVERSITY does not:
admitted distinct keys peak at round 6 (823) then fall monotonically (750, 686).
C6 also holds the most distinct rich programs on the held-out benchmark (216,
against C5's 213, C7's 199, and C8's 205), so selecting it is defensible on both
measures --- but the benchmark is a single 256-sample draw per checkpoint with no
variance estimate: C5's 213 sits within a hair of C6's 216, and the C7-vs-C8
order (205 > 199) inverts the training curve's (750 > 686), so around the peak
the draws disagree within sampling noise. The
load-bearing evidence is therefore the training-curve decline, not the
benchmark: past round 6 the policy trades breadth for reward --- mild
mode-narrowing that the anti-collapse guards of \S{}5.5 bound but do not abolish ---
so the method has an optimal stop visible in the distinct-key curve rather than
the rich-rate. What this replication establishes is bounded: verifier-only
fine-tuning is not a ONE-grammar artifact --- run unchanged on a language with no
data, no market, and no trades, only checked arithmetic and bounded loops, it
takes a small model from 3.5\% to \textasciitilde{}95\% rich generation with \textasciitilde{}100\% corpus-novelty.
It does not establish generality past the confounds both arms share --- same kit
and author (as for the construction claim, \S{}4.5) and, specific to the fine-tune,
the same base \code{Qwen3-0.6B}, trainer, and validity-ladder reward shape; generality
across models in particular is untested. What it also
does not establish is that every rich program is INTERESTING; the RICH rung
certifies well-formed, terminating,
and varied output, and selecting genuinely useful programs from this competent
generator stays a downstream concern, exactly as edge does for stratlite. As in
\S{}5.6 the MODEL does not reproduce, but the SCORING does, from the committed
pools.

\subsection*{5.8 Transfer: solving held-out problems, and what self-play optimizes}

The abstract flags the gap honestly: \S{}5.6--5.7 measure generation VALIDITY,
where tempo-x402 measured task-solving pass@1. This section closes that gap
with a held-out problem-solving benchmark on \code{prooflite}
(\code{experiment/proofbench/problems/heldout.jsonl}): 30 specs in three tiers
(9 easy / 9 medium / 12 hard --- the hard tier needs real algorithms: primality,
popcount, digit-sum, Collatz length), each with a verified reference solution.
A candidate SOLVES a problem iff its output exactly equals the reference's;
8 samples per problem per model, and the pools are committed
(\code{results/solve\_\{base,cinit,c6\}.jsonl}), so `p6 solve problems/heldout.jsonl
results/solve\_<m>.jsonl` reproduces every number. Three arms separate the
recipe's ingredients: \textbf{base} \code{Qwen3-0.6B}; \textbf{Cinit}, the cold-start
checkpoint --- supervised fine-tuning on the 174 human corpus programs and
nothing else, i.e. the plain-SFT baseline; and \textbf{C6}, Cinit plus six rounds of
verifier-only self-play --- the checkpoint \S{}5.7 selects on the generation
benchmark. (Cinit's own generation row, same 256-sample protocol: 83.2\% parse,
48.8\% RICH, 124 distinct rich keys --- between base and the self-play curve, as
expected.)

\begingroup\small
\begin{longtable}{p{0.129\linewidth} p{0.129\linewidth} p{0.129\linewidth} p{0.129\linewidth} p{0.129\linewidth} p{0.129\linewidth} p{0.129\linewidth}}
\toprule
\textbf{model} & \textbf{RICH generation} & \textbf{solve pass@8} & \textbf{easy} & \textbf{medium} & \textbf{hard} & \textbf{pad} \\
\midrule\endhead
base & 3.5\% & 20.0\% (6/30) & 3/9 & 2/9 & 1/12 & 1 \\
Cinit (plain SFT) & 48.8\% & \textbf{80.0\%} (24/30) & 9/9 & 8/9 & 7/12 & 1 \\
C6 (+ self-play) & \textbf{94.5\%} & 43.3\% (13/30) & 6/9 & 5/9 & 2/12 & 8 \\
\bottomrule
\end{longtable}\endgroup

Two findings, one expected and one not --- and the second is the sharper.

\textbf{Transfer is real.} Training on verifier-selected programs --- never on
problem/solution pairs --- takes pass@8 from 20\% to 80\%, including 7 of the 12
hard discriminators. Base solves a few easy problems from surface familiarity
alone (the C-like syntax parses often enough for counting loops to land), so
the floor is not zero here; the lift is what carries the tempo-x402 shape onto
this instrument, now in its original task-solving units.

\textbf{Self-play optimizes the verifier, and the verifier is not the task.} The
checkpoint that wins generation (C6, 94.5\% RICH against Cinit's 48.8\%) loses
nearly half of Cinit's solving (43.3\% against 80.0\%). The \code{pad} column is the
mechanism: \code{p6 solve} separately reports candidates whose output has the
reference answer as a strict prefix followed by extra output --- the right
algorithm, then spurious prints. Eight of C6's seventeen misses are exactly
that (against one each for base and Cinit): a correct \code{sum\_1\_100} loop
followed by a dozen filler \code{print} statements is the RICH rung ($\ge$3 distinct
output lines over $\ge$30 fuel) speaking, distilled by self-play into an
unconditional output habit that overrides the spec. The localization is exact.
On the 12 problems whose correct output is itself RICH-shaped ($\ge$3 distinct
lines), C6 matches Cinit --- 9/12 versus 10/12; on the 18 problems whose correct
output is short --- where the training reward actively disprefers the correct
shape --- Cinit solves 14/18 while C6 solves 4/18 and pads 8 of the remaining
\begin{enumerate}[leftmargin=*]
\item Where reward and task agree, self-play costs nothing measurable; where
\end{enumerate}
they conflict, the policy obeys its internalized reward, not the prompt.

The two capabilities therefore move in OPPOSITE directions under self-play ---
generation validity 3.5 $\rightarrow$ 48.8 $\rightarrow$ 94.5, spec-conditioned solving 20 $\rightarrow$ 80 $\rightarrow$ 43 ---
and the diversity narrowing of \S{}6.6 has a sharper companion: the policy does
not merely narrow, it carries the reward's preferences into contexts the
reward never saw. This is the Goodhart argument of \S{}6.3 measured at the policy
level, and it revises the recipe's reading. Cold-start SFT buys the language
AND the ability to use it on demand; self-play buys the verifier's rungs,
specifically --- run it only while its rungs and the downstream task agree, and
read the pad diagnostic as the early warning. The caveats are the section's
own: pass@8 over 30 problems is a coarse single-draw instrument; one base
model, one language; and the prefix-shaped \code{pad} count is a lower bound on
"right answer, wrong shape" (answer-inside-padding is not counted). Solving
was also never in the training objective for EITHER checkpoint --- the surprise
is not that C6 solves worse than a solver would, but that plain SFT solves
this well and self-play then gives half of it back.

\subsection*{5.9 PENDING (permanently keyless): the frozen-model A/B arm}

One experiment on the instrument has complete plumbing and a committed protocol
but has not been run, and cannot be: the frozen-model A/B arm. The \S{}5 selection
comparison --- arm V (verified pick: among survivors, best train PnL, re-scored on
held-out) versus arm U1 (the naive user: ask once, ship candidate 0 unverified)
--- consumes a batch generated by a frozen API model (\code{claude-opus-4-8}, no
temperature/seed, so the batch is recorded not reproduced;
\code{experiment/src/api.rs:21}). Both arms consume the identical pool, so
best-of-N optimism is symmetric and cancels in the paired difference. This arm
has \textbf{no API key and will not get one} (\code{experiment/src/api.rs:69}); the
selection logic and its scoring are shipped and tested (\code{score::pick\_verified},
\code{score::score\_heldout}, \code{cd experiment \&\& cargo test}), and the \S{}5.2 agent
corpus is the key-free stand-in for the generation step.

\begin{itemize}[leftmargin=*]
\item Generate: \code{cd experiment \&\& cargo run -- submit <n> batch.json} then
\end{itemize}
  \code{cargo run -- poll <batch\_id> raw.jsonl} (needs \code{ANTHROPIC\_API\_KEY})
\begin{itemize}[leftmargin=*]
\item Score the two arms (pure, no network):
\end{itemize}
  \code{cd experiment \&\& cargo run -q -- score raw.jsonl data/BTCUSDT-1h-2024-01.csv}

The reproducibility split is exact and stated once: the deterministic verifier
plus committed artifacts reproduce every number in \S{}5.2--5.8 from a command ---
including both fine-tunes' scoring (stratlite's held-out gate-clear from committed
candles, prooflite's RICH-rate, novelty, and solve pass@8 from committed pools
and corpus); the
generation itself --- agents, the frozen API model, or either fine-tune --- does not,
and is committed or recorded rather than regenerated.

\bigskip\hrule\bigskip

\section*{6. Limits and negative results}

The thesis of this paper is that smallness buys mechanical guarantees. This
section reports where those guarantees did not bind, where they can be gamed,
and what they provably cannot certify. Each finding is measured, not argued, and
every figure below reproduces from a command in the repository.

\subsection*{6.1 On this task the fuel bound was free --- which means it was not tested}

The termination guarantee is the kit's headline: a \code{stratlite} program provably
halts within its per-bar fuel, so computing its backtest --- or its training
reward --- can never hang. On the corpus we can run, the guarantee never came
close to binding. Across the 132 survivors of the 134-program corpus, per-bar
fuel consumption ranges from a minimum of 16 to a median of 55 to a maximum of
186, against a default cap of 25,000 fuel per bar (\S{}5.3; the field is emitted by
`cd experiment \&\& cargo run -q -- reward corpus/seed.jsonl
data/BTCUSDT-1h-2024-01.csv\code{ over }class:"ok"` rows; the cap is
\code{Limits::default().fuel\_per\_bar} at \code{crates/stratlite/src/lib.rs:137}). The most
expensive strategy any of six diverse agent families produced spends 0.74\% of
the budget.

This is a negative result about our own evidence, and it must be stated
precisely. It does \textbf{not} falsify the termination guarantee: the guarantee is a
proof about the evaluator, not a claim about the corpus, and it holds whether or
not any program approaches the cap. What it falsifies is any claim that this
corpus \emph{exercised} the bound. As a proof the fuel cap is sound; as a
discriminator --- a filter that rejects programs --- it did zero work here, so its
discriminating power is untested. The honest artifact is therefore the full
distribution (min 16 / median 55 / max 186), not the count of fuel-exhaustion
faults, because a bare "0 programs exhausted fuel" cannot distinguish "the bound
did its job" from "the bound was decoration." A regime that actually pressures
the cap (deeper indicator windows, adversarial rather than cooperative
generation) remains future work; until then the strongest claim the data
supports is that fuel is cheap insurance whose premium, on cooperative inputs,
is zero. That is task-specific, though: the N=2 \code{prooflite} fine-tune (\S{}5.7)
pushed measured fuel use, among its \code{ok}-rung programs, to a maximum of 55,756
of the 100,000-fuel default --- 55.8\% of budget, against stratlite's 0.74\%, some
75$\times$ the budget fraction and far nearer the cap. On a compute language with real loops the
bound came far closer to binding, so "the fuel bound was free" is a fact about
the stratlite backtest, not about the kit.

\subsection*{6.2 The single-month benchmark has no out-of-sample teeth}

The \S{}5.4 selection experiment needs a held-out window that is genuinely harder
than the training window; otherwise a "survivor lift" measures grammar-learning,
not generalization. On the one month of hourly BTCUSDT candles we ship, the
held-out window is not harder. Among the programs that compile, the gate-clear
rate is 98.5\% on train and 97.0\% on held-out --- a gap of 1.5 points --- and
per-style only mean-reversion shows any spread (86\% held-out versus 100\% for
four of the other five families; `cd experiment \&\& cargo run -q -- eval
corpus/seed.jsonl data/BTCUSDT-1h-2024-01.csv`, full table in \S{}5.4). The
aggregate gap is near noise.

This was predicted before it was measured. The compile rung is
data-independent --- whether a program parses does not depend on the candles --- so a
raw held-out survivor rate can rise entirely on grammar competence that shows
identically on train. The eval metric is therefore deliberately \emph{conditional}
(gate-clear among compilers) rather than raw, and the pre-registered discipline
(\code{experiment/M6.md}) requires proving the benchmark has out-of-sample teeth
before any lift on it is claimed. The measurement says it does not, on one month
of one asset. The consequence shaped the M6 fine-tune benchmark (\S{}5.6): rather
than score the fine-tune on this single month, we built the wider regime split
this section demands --- a chronologically distant BTC window under a five-month
embargo and a second asset, ETH, never trained on. On that split C7's held-out
gate-clear generalizes (95.7\% / 96.5\% / 96.1\%), yet the train-minus-held-out gap
stays near zero (2.3 / 1.2 / 1.6 points). The wider split thus confirms that
what generalizes is the \emph{language} --- grammar competence is regime- and
asset-independent --- and still shows no out-of-sample edge teeth, because the gap
that would carry an edge claim is exactly the gap that stays near zero. We
therefore report a competence lift off a measured-zero floor (\S{}5.6), not an
out-of-sample edge result.

\subsection*{6.3 Goodhart risk: hard caps push complexity into the seams}

The constitution's LOC caps ($\le$2,000 per crate, $\le$25,000 repo) are mechanical and
CI-enforced, and mechanical targets invite Goodhart's law: complexity that
cannot live inside a capped crate migrates to wherever the counter cannot see
it. The natural escape hatch is the experiment harness, which sits outside the
kit's Cargo workspace precisely because it is allowed dependencies the kit
forbids. If the caps counted only kit crates, the harness would be an unmetered
sink into which any inconvenient logic could be pushed while every per-crate
number stayed green.

The mitigation is to extend the counter across the seam rather than trust a
boundary. \code{scripts/caps.sh} meters \code{experiment/src} at a 1,500-LOC cap
(currently 1,489), the Python trainer at an 800-LOC cap (currently 639), and the
N=2 reward tool \code{experiment/proofbench/src} at a 1,500-LOC cap (currently 354),
alongside the kit's per-crate 2,000 and repo 8,214 / 25,000
(\code{bash scripts/caps.sh}). The harness cap sits 11 lines below its ceiling, which
is itself a signal: the counter is close enough to bite, so it is not
decorative. This does not eliminate Goodhart pressure --- a determined author could
still relocate complexity into a text corpus, a data file, or prose the LOC
counter ignores --- but it closes the one seam the architecture most invites, and
it makes the cost of the seam visible in the same report as the caps it
protects. The general limit stands: a mechanical cap disciplines the thing it
counts and nothing else, and every uncounted surface is a place the discipline
does not reach.

\subsection*{6.4 The seam-tax question is open; M5 has not answered it}

The whole kit rests on a claim reality has not yet tested: that extracting the
kernel three parent languages hand-rolled is a net simplification for a real
consumer, not a seam-tax that makes every language pay an integration cost
larger than the duplication it removes. The four milestones on the kit
(\code{prooflite}, \code{caplite}, the two emitters, \code{stratlite}+\code{backtestlite}) show the
kit \emph{composes}, but every one of them was built to consume the kit --- they cannot
be disinterested witnesses. The two emitters are the sharpest instance: 2,170
LOC built and tested with no consumer anywhere in this repo, whose named
consumer is precisely this pending re-homing, so their reuse payoff is argued
from the parents rather than shown. The designed test is M5: re-home the parent
\code{bashlite} onto the kit inside \code{localharness} and require the migration to shed
net LOC there, or it does not ship. That milestone has not been run. Until it
is, the load-bearing claim of the paper --- that the kernel carries its weight for
a consumer that did not exist to justify it --- is supported by construction
convenience and argument, not by an independent measurement. We name this as the
single largest unverified claim in the work, not as a promise that it will come
back positive.

\subsection*{6.5 Reward hacks in our own reward: the adversarial frame is necessary, not rhetorical}

M6 uses the verifier as a training reward, and building it surfaced concrete
ways a model could farm reward without producing a strategy (\S{}5.5 enumerates the
five hacks and their guards). What matters for the limits discussion is that
these were found in our own oracle, which is the evidence that the adversarial
framing earns its keep rather than decorating the paper. Two are closed and
tested with no GPU --- the empty-rollout hack (an empty source now scores a
compile-class zero, \code{the\_empty\_rollout\_hack\_is\_closed}, one of eight tests
matching \code{cd experiment \&\& cargo test}) and the gate-rung hack (\code{lookback 4;}
collects 2/3 for the absence of a strategy, so admission keeps Ok-rung survivors
only, \code{test\_only\_ok\_rung\_is\_admitted}, in the six-test stdlib-only trainer suite
\code{cd experiment/train \&\& python3 test\_select.py}). Related hacks ---
\code{equity\_hash} diversity gamed by a one-tick perturbation, mode collapse onto one
template --- drove the switch to a source-canonical \code{novelty\_key} with a
documented honest limit (it catches comment and format clones, not
constant-perturbation or semantic clones).

The general finding: a reward oracle is an adversary's target, and the failures
were not exotic --- they were the cheapest programs in the language. That they
existed in a verifier we built deliberately, and were found only by attacking
it, is the concrete case for treating verification as physics across an
adversarial boundary rather than as a checklist. It also bounds the claim: the
guards close the hacks we found. We do not claim to have enumerated the reward's
attack surface, and AST-level canonicalization --- the refinement that would close
the semantic-clone class --- is unbuilt.

\subsection*{6.6 The generator narrows past its diversity peak, and two untested edges}

The M6 fine-tune carries a negative result of its own, visible as a pattern
rather than a fluke only because N=2 showed it twice. In both arms the admitted
programs' distinct-key count rises, peaks, then declines while raw validity keeps
climbing: \code{prooflite} peaks at round 6 (823 distinct keys) and falls to 686 by
round 8 (\S{}5.7); \code{stratlite} peaks at round 4 (694) and falls to 534 by round 7
(\S{}5.6), the steeper drop of the two. Past the peak the policy trades breadth for
reward --- concentrating on a narrower band of easy, high-scoring programs --- a mild
mode-narrowing the anti-collapse guards of \S{}5.5 bound but do not abolish. So
"train longer" is not free: the generator has an optimal stop, and it lives in
the distinct-key curve, not the rich-rate the runs actually early-stopped on.
\code{prooflite}'s C6 was selected at its diversity peak; \code{stratlite}'s C7 was
selected on rich-rate saturation and therefore sits past its own --- its headline
validity numbers are unaffected (validity is monotone), but its generator is
less varied than an R4 checkpoint's would be. And narrowing is the milder half
of what self-play does to the policy: \S{}5.8 measures the sharper half --- the
reward's output shape persisting as an unconditional habit that overrides
held-out specs, halving the plain-SFT checkpoint's problem-solving where spec
and reward shape conflict.

Two further edges of the fine-tune stay untested, and belong here rather than
only inline in \S{}5.7. First, cross-model generality: both arms fine-tune the same
base model (\code{Qwen3-0.6B}), so the lift could in principle reflect that model's
inductive biases rather than the recipe --- nothing here varies the model. Second,
the novelty control (\S{}5.7) is measured only against the human cold-start corpus,
the model's sole external data; it rules out memorizing those 174 examples, not
reproduction from the far larger set of the model's own self-play programs, which
was not separately checked. Neither undoes the results, but a skeptic reading the
limits section should find both stated, not buried.

\subsection*{6.7 What smallness cannot buy}

Smallness buys decidable \emph{form}: this program halts, touches only these
capabilities, emits at most this many bytes, is active enough to be a strategy.
It does not buy \emph{semantic correctness beyond the checked properties}, and on
this corpus the gap is stark. All 132 survivors clear every mechanical gate ---
they compile, they provably halt, they respect the effect bound, they trade
actively enough to pass the gate. Of those 132, 114 lose money on the very
window they were evaluated against; only 18 are profitable on train (the
\code{train\_pnl} field over \code{class:"ok"} rows of the \code{reward} command in \S{}6.1). The
verifier certifies that a program is a well-formed, terminating, active
strategy. It says nothing about whether that strategy is a \emph{good} one, and by
construction it must not: train PnL is deliberately excluded from the reward
(\S{}5.5), because rewarding it teaches curve-fitting and makes any held-out
comparison circular.

This is the boundary of the entire approach, stated plainly. A purpose-sized
language makes a chosen set of properties mechanical and complete; it makes no
other property true. Edge, correctness of intent, robustness to a regime it
never saw --- these live outside the checked set and stay the job of selection,
out-of-sample evaluation, and human judgment. The right reading of "132
survivors, 114 unprofitable" is not that the verifier failed but that it
succeeded at exactly what it claims and nothing more: it separated well-formed
strategies from noise, and left the question of which well-formed strategy is
worth running entirely open. Where the properties a task needs fall outside any
small decidable set --- open-ended correctness, rich effects, behavior that only
tests can pin down --- a general-purpose language plus a test suite remains the
better tool, and this kit makes no claim on that ground.

\bigskip\hrule\bigskip

\section*{Appendix: Reproducibility}

Every number in this paper is reproducible by a command in the repository, with
one honest exception stated once and everywhere it applies: the \emph{generation} of
programs (by agents, a frozen API model, or a fine-tune) does not reproduce and
is committed or recorded, not regenerated; the deterministic verifier plus the
committed artifacts reproduce every figure below. All commands run from the repo
root unless a \code{cd} is shown.

\subsection*{Kit structure, size, and caps --- \code{bash scripts/caps.sh}}

\begingroup\small
\begin{longtable}{p{0.450\linewidth} p{0.450\linewidth}}
\toprule
\textbf{Number} & \textbf{Command} \\
\midrule\endhead
repo total 8,214 LOC (cap 25,000) & \code{bash scripts/caps.sh} \\
\code{diaglite} 252 LOC & \code{bash scripts/caps.sh} \\
\code{lexlite} 290 LOC & \code{bash scripts/caps.sh} \\
\code{parselite} 238 LOC & \code{bash scripts/caps.sh} \\
\code{fuellite} 180 LOC & \code{bash scripts/caps.sh} \\
\code{caplite} 521 LOC & \code{bash scripts/caps.sh} \\
\code{evmlite} 1,400 LOC & \code{bash scripts/caps.sh} \\
\code{modlite} 770 LOC & \code{bash scripts/caps.sh} \\
\code{prooflite} 1,955 LOC (cap 2,000) & \code{bash scripts/caps.sh} \\
\code{stratlite} 1,893 LOC & \code{bash scripts/caps.sh} \\
\code{backtestlite} 669 LOC & \code{bash scripts/caps.sh} \\
\code{stratlite} + \code{backtestlite} = 2,562 LOC & \code{bash scripts/caps.sh} (1,893 + 669) \\
kernel \code{diaglite}+\code{lexlite}+\code{parselite}+\code{fuellite} = 960 LOC & \code{bash scripts/caps.sh} (252+290+238+180) \\
seven kit crates = 3,651 LOC & \code{bash scripts/caps.sh} (sum) \\
emitters combined 2,170 LOC (built, unconsumed in repo) & \code{bash scripts/caps.sh} (1,400 + 770) \\
\code{experiment/src} 1,489 LOC (cap 1,500); harness 11 lines below ceiling & \code{bash scripts/caps.sh} \\
\code{experiment/train} 639 LOC (cap 800) & \code{bash scripts/caps.sh} \\
\code{experiment/proofbench/src} 354 LOC (cap 1,500) --- the N=2 reward tool & \code{bash scripts/caps.sh} \\
\code{CLAUDE.md} 7,963 chars (cap 8,000) & \code{bash scripts/caps.sh} \\
per-crate cap 2,000, repo cap 25,000, \code{CLAUDE.md} cap 8,000 & \code{scripts/caps.sh} (CRATE\_CAP/REPO\_CAP/CLAUDE\_CAP) \\
zero external dependencies, all 11 crates & inspect each \code{crates/*/Cargo.toml} \code{[dependencies]} (workspace-internal entries only) \\
\code{wasm32-unknown-unknown} target green (exit 0) & \code{cargo check --target wasm32-unknown-unknown} \\
\bottomrule
\end{longtable}\endgroup

\subsection*{Test counts and pinned invariants --- \code{cargo test -p <crate>}}

\begingroup\small
\begin{longtable}{p{0.450\linewidth} p{0.450\linewidth}}
\toprule
\textbf{Number} & \textbf{Command} \\
\midrule\endhead
\code{diaglite} 7 tests & \code{cargo test -p diaglite} \\
\code{lexlite} 7 tests & \code{cargo test -p lexlite} \\
\code{parselite} 4 tests & \code{cargo test -p parselite} \\
\code{fuellite} 4 tests & \code{cargo test -p fuellite} \\
\code{caplite} 10 tests & \code{cargo test -p caplite} \\
\code{evmlite} 17 tests & \code{cargo test -p evmlite} \\
\code{modlite} 8 tests + 1 doctest & \code{cargo test -p modlite} \\
\code{litelite} facade 1 test & \code{cargo test -p litelite} \\
58 unit tests + 1 doctest across the kit & sum of the eight kit crates above \\
\code{prooflite} 31 unit tests + 2 doctests & \code{cargo test -p prooflite} \\
\code{stratlite} 10 unit tests + 1 doctest (11 pass) & \code{cargo test -p stratlite} \\
\code{backtestlite} 6 unit tests + 1 doctest (7 pass) & \code{cargo test -p backtestlite} \\
cost model exact: \code{print 1;}=2, \code{let x = 1 + 2;}=4, \code{repeat 3 \{ print 0; \}}=11 & \code{cargo test -p prooflite} (\code{the\_cost\_model\_is\_exact}) \\
operator chain of 50 evaluates, 500 trips \code{E0102}, 5,000 parse-and-drop safe & \code{cargo test -p prooflite} (\code{operator\_chains\_count\_toward\_the\_depth\_cap}) \\
\code{caps()} fetched exactly once per run & \code{cargo test -p prooflite} (\code{the\_validated\_table\_snapshot\_drives\_the\_whole\_run}) \\
\code{equity\_hash} FNV-1a-64, determinism pinned & \code{cargo test -p backtestlite} (\code{determinism\_is\_exact\_and\_hashable}) \\
no-look-ahead prefix-invariance & \code{cargo test -p stratlite} (\code{no\_lookahead\_prefix\_invariance}) \\
\bottomrule
\end{longtable}\endgroup

\subsection*{Source constants and signatures --- \code{Read crates/... / experiment/...}}

\begingroup\small
\begin{longtable}{p{0.450\linewidth} p{0.450\linewidth}}
\toprule
\textbf{Number} & \textbf{Command} \\
\midrule\endhead
\code{DEFAULT\_MAX\_DEPTH = 96} & \code{crates/parselite/src/lib.rs:25} \\
\code{prooflite} default fuel 100,000 / output 64 KiB & \code{crates/prooflite/src/lib.rs:247-248} (\code{Limits::default}) \\
\code{stratlite} \code{fuel\_per\_bar = 25\_000} & \code{crates/stratlite/src/lib.rs:137} \\
\code{verify()} $\rightarrow$ \code{Result<(Strategy, Report), Reject>} & \code{crates/backtestlite/src/lib.rs:199} \\
\code{enum Reject \{ Compile \textbar{} Run \textbar{} Gate \}} & \code{crates/backtestlite/src/lib.rs:124} \\
gate default \code{min\_trades = 4}, \code{min\_bars\_evaluated = 16} & \code{crates/backtestlite/src/lib.rs:89} \\
\code{stratlite::REFERENCE} generation prompt (a \code{const}) & \code{crates/stratlite/src/lib.rs:78} \\
M6 default model \code{Qwen/Qwen3-0.6B} (trainer config) & \code{experiment/train/train.py:55} \\
frozen model \code{claude-opus-4-8} & \code{experiment/src/api.rs:21} \\
A/B arm requires \code{ANTHROPIC\_API\_KEY} (permanently keyless) & \code{experiment/src/api.rs:69} \\
\bottomrule
\end{longtable}\endgroup

\subsection*{Construction cost --- git history and \code{GENESIS.md}}

\begingroup\small
\begin{longtable}{p{0.450\linewidth} p{0.450\linewidth}}
\toprule
\textbf{Number} & \textbf{Command} \\
\midrule\endhead
genesis \code{a557fef} 2026-07-14 22:10:31 & \code{git show -s --format='\%h \%ad' --date=format:'\%Y-\%m-\%d \%H:\%M:\%S' a557fef} \\
\code{prooflite} land \code{07c2877} 00:12:30, review \code{9ac20b3} 00:40:37 (28 min); M3 done \code{3d5918e} 02:25:10; \code{stratlite}/\code{backtestlite} land \code{42667a6} 02:59:24, review \code{1524eb3} 03:34:19 (35 min); M4 span 1 h 09 m; end-to-end 5 h 24 m & \code{git log --format='\%h\textbar{}\%ad\textbar{}\%s' --date=format:'\%Y-\%m-\%d \%H:\%M:\%S' a557fef..1524eb3} \\
parents: \code{rustlite} 8.0K LOC/99 tests, \code{soliditylite} 7.6K/159, \code{bashlite} 3.1K/64 (external, not repo-reproducible) & \code{GENESIS.md}, "The lineage" \\
\code{bashlite} shipped with no parser depth guard & \code{GENESIS.md}, "What was ported at genesis" \\
\code{prooflite} review 16 raw / 13 confirmed / 2 crash-grade (M1); M2 16 confirmed; \code{stratlite}+\code{backtestlite} 14 confirmed (M4) & \code{GENESIS.md}, "Post-genesis lessons" \\
roadmap budgeted \textasciitilde{}a week per language & \code{GENESIS.md}, "The three questions only reality can answer" \\
predecessor: 0.5B model 1.5\% $\rightarrow$ 16.4\% pass@1, 201-problem benchmark (tempo-x402, prior work, not recomputed here) & \code{GENESIS.md}, "The lineage"; \code{paper/OUTLINE.md} \\
\bottomrule
\end{longtable}\endgroup

\subsection*{Experiment run --- corpus + verifier (all deterministic)}

\begingroup\small
\begin{longtable}{p{0.450\linewidth} p{0.450\linewidth}}
\toprule
\textbf{Number} & \textbf{Command} \\
\midrule\endhead
corpus 134 programs; trend 23 / breakout 23 / mean-reversion 22 / momentum 22 / stateful 22 / combo 22 & \code{wc -l experiment/corpus/seed.jsonl}; \code{Counter} over the \code{style} field \\
744 candles; split train 0..446 / held-out 446..744; costs fee 2196 / slip 439 ticks (5 bps / 1 bps, train-only); drift 0.943x & \code{cd experiment \&\& cargo run -q -- data data/BTCUSDT-1h-2024-01.csv} \\
Reject histogram 0 compile / 0 run / 2 gate / 132 ok = 100\% compile, 132/134 = 98.5\% survivor & \code{cd experiment \&\& cargo run -q -- reward corpus/seed.jsonl data/BTCUSDT-1h-2024-01.csv} (tally \code{class}) \\
fuel over survivors min 16 / median 55 / max 186 vs 25,000 cap = 0.74\% at max & \code{cd experiment \&\& cargo run -q -- reward corpus/seed.jsonl data/BTCUSDT-1h-2024-01.csv} (\code{fuel} over \code{class:"ok"}) \\
134 distinct \code{novelty\_key} over 134 programs & \code{cd experiment \&\& cargo run -q -- reward corpus/seed.jsonl data/BTCUSDT-1h-2024-01.csv} (distinct \code{nkey}) \\
114 of 132 survivors negative \code{train\_pnl}, 18 positive & \code{cd experiment \&\& cargo run -q -- reward corpus/seed.jsonl data/BTCUSDT-1h-2024-01.csv} (\code{train\_pnl} over \code{class:"ok"}) \\
conditional eval: compile 100\%, gate-clear train 98.5\% vs held-out 97.0\%, gap 1.5 points & \code{cd experiment \&\& cargo run -q -- eval corpus/seed.jsonl data/BTCUSDT-1h-2024-01.csv} \\
per-style held-out gate-clear: breakout 100\% (23/23), combo 100\% (22/22), momentum 100\% (22/22), stateful 100\% (22/22), trend 96\% (22/23), mean-reversion 86\% (19/22) & \code{cd experiment \&\& cargo run -q -- eval corpus/seed.jsonl data/BTCUSDT-1h-2024-01.csv} \\
reward ladder Compile 0 / Run 1/3 / Gate 2/3 / Ok 1; \code{value == ladder(class)}; anti-hacking guards; 23 tests pass (8 match \code{reward}) & \code{cd experiment \&\& cargo test} \\
Python admission guards (Ok-rung only, \code{novelty\_key} dedup, per-style cap) stdlib-only; 6 pass & \code{cd experiment/train \&\& python3 test\_select.py} \\
\bottomrule
\end{longtable}\endgroup

\subsection*{Fine-tune benchmark (M6) --- reproducible scoring; model not reproducible}

The MODEL does not reproduce (stochastic sampling; a fine-tune is not
bit-identical across hardware). The SCORING does: the sample pools and candles
are committed, and each row below reproduces from a command. \code{target/} is
git-ignored, so build the verifiers first on a clean clone: `cd experiment \&\&
cargo build --release\code{ (s5) and }cd experiment/proofbench \&\& cargo build
--release` (p6).

\begingroup\small
\begin{longtable}{p{0.450\linewidth} p{0.450\linewidth}}
\toprule
\textbf{Number} & \textbf{Command} \\
\midrule\endhead
base compile / gate-clear 0.0\% / 0.0\% on all three windows & \code{cd experiment \&\& ./target/release/s5 eval results/base.jsonl data/<window>.csv} \\
C7 BTC Jan 2024 (train window) 100.0\% / 95.7\% & \code{cd experiment \&\& ./target/release/s5 eval results/c7.jsonl data/BTCUSDT-1h-2024-01.csv} \\
C7 BTC Jun 2024 (5-month embargo) 100.0\% / 96.5\% & \code{cd experiment \&\& ./target/release/s5 eval results/c7.jsonl data/BTCUSDT-1h-2024-06.csv} \\
C7 ETH Jun 2024 (cross-asset) 100.0\% / 96.1\% & \code{cd experiment \&\& ./target/release/s5 eval results/c7.jsonl data/ETHUSDT-1h-2024-06.csv} \\
train-minus-held-out gap 2.3 / 1.2 / 1.6 points (BTC Jan / BTC Jun / ETH Jun) & same \code{s5 eval} commands (TRAIN vs HELD-OUT line) \\
per-style held-out gate-clear near-uniform 88--100\% on the cross-asset window & \code{cd experiment \&\& ./target/release/s5 eval results/c7.jsonl data/ETHUSDT-1h-2024-06.csv} \\
training curve full-survivor cold-start $\rightarrow$ R0 39.0\% $\rightarrow$ R7 98.2\% (8 rounds); \code{distinct\_nkeys} peaks R4 (694) then narrows to 534 (R7) & \code{experiment/results/train\_curve.log} \\
pools 256 samples each (32/style $\times$ 8 styles); full benchmark output & \code{experiment/results/\{base,c7\}.jsonl}; \code{experiment/results/README.md}, \code{experiment/results/benchmark.txt} \\
C7 novelty 99.6\% (250/251 gate-clearers $\notin$ corpus) & \code{cd experiment \&\& ./target/release/s5 reward results/c7.jsonl data/BTCUSDT-1h-2024-01.csv} (compare \code{nkey} field vs \code{s5 reward corpus/seed.jsonl ...}) \\
N=2 prooflite: base RICH 3.5\% (9/256), C6 RICH 94.5\% & \code{cd experiment/proofbench \&\& ./target/release/p6 eval results/base.jsonl} then \code{... eval results/c6.jsonl} (one pool per run) \\
N=2 prooflite: novel rich \textasciitilde{}100\% ($\notin$ 174-program corpus); C6 216 distinct rich & \code{cd experiment/proofbench \&\& ./target/release/p6 novelty results/c6.jsonl corpus/seed.jsonl} \\
N=2 prooflite training curve rich-rate R0 43.8\% $\rightarrow$ R8 96.2\% (9 rounds); \code{distinct\_nkeys} peaks R6 (823) then narrows to 686 (R8) & \code{experiment/proofbench/results/train\_curve.log} \\
N=2 prooflite pools 256 each; full output & \code{experiment/proofbench/results/\{base,c6,c7,c8\}.jsonl}; \code{experiment/proofbench/results/README.md}, \code{.../benchmark.txt} \\
\bottomrule
\end{longtable}\endgroup

\subsection*{PENDING slot --- instrument shipped, run not performed (permanently keyless)}

\begingroup\small
\begin{longtable}{p{0.450\linewidth} p{0.450\linewidth}}
\toprule
\textbf{Slot} & \textbf{Command} \\
\midrule\endhead
A/B arm: generate & \code{cd experiment \&\& cargo run -- submit <n> batch.json} then \code{cargo run -- poll <batch\_id> raw.jsonl} (needs \code{ANTHROPIC\_API\_KEY}) \\
A/B arm: score the two arms (pure, no network) & \code{cd experiment \&\& cargo run -q -- score raw.jsonl data/BTCUSDT-1h-2024-01.csv} \\
\bottomrule
\end{longtable}\endgroup
\end{document}
